---
title: "Common Data Manipulations in R"
author: "Andrew MacDonald"
date: '2014-11-07'
output: 
    html_document:
      keep_md: TRUE
---

# Simple data manipulations in R

Many years ago, I was introduced to R by [Cam Webb ](http://camwebb.info/).  At the time, his website contained a list of common data manipulations (original [here](http://camwebb.info/blog/2014-04-29/)).  This list dated from Cam's early experience with R, and contained the R-help mailing list responses to a series of data manipulations.  For a long time, I kept this file as a handy reference.  I printed it out.  I recommended it to friends.

Now I have been using R for years, and the state of the art has advanced considerably.  Particulary, [Hadley Wickham's](https://github.com/hadley) `reshape2` and `dplyr` packages have transformed the way most useRs manipulate their data.  I decided that it would be interesting to revisit my favourite resource and try my hand at solving these problems with tools from these two packages.


```{r message=FALSE}
library(reshape2)
library(dplyr)
library(knitr)
```





## GROUP

Turn this table (A):


| c1 | c2 | c3 |
|----|----|----|
| A  | a  |  1 |
| A  | a  |  3 |
| A  | a  |  1 |
| A  | b  |  1 |
| A  | b  |  2 |
| B  | c  |  2 |
| B  | d  |  1 |


into this (B):

| Group1 | Group2 | Nrows | SumOfCol3 |
|--------|--------|-------|-----------|
| A      | a      |     3 |         5 |
| A      | b      |     2 |         3 |
| B      | c      |     1 |         2 |
| B      | d      |     1 |         1 |



```{r results='asis'}
A <- data.frame(
       c1 = c('A', 'A', 'A', 'A', 'A', 'B', 'B'),
       c2 = c('a', 'a', 'a', 'b', 'b', 'c', 'd'),
       c3 = c(1, 3, 1, 1, 2, 2, 1))

B <- A %>%
  group_by(c1,c2) %>%
  summarize(Nrows=n(),
            SumOfCol3=sum(c3))
B %>%
	ungroup %>%
	kable
```


## SPLIT

Turn col3 of the second table (B) into this (C):

| row | a | b | c | d |
|-----|---|---|---|---|
| A   | 3 | 2 | . | . |
| B   | . | . | 1 | 1 |


```{r results='asis'}
C <- dcast(B,c1~c2,value.var="Nrows")
kable(C)
```


Many original responders suggested the use of `table`, referring to the original dataset:


```{r results='asis'}
C_alt<-with(A,table(c1,c2))
kable(C_alt)
```


although that solution is not ["tidy" in the Hadlian sense](http://vita.had.co.nz/papers/tidy-data.pdf) -- i.e., it does not return a `data.frame`, but rather a `table` object.  You can obtain a data.frame with `dcast` directly:


```{r results='asis'}
C_alt2<-dcast(A,c1~c2,value.var="c3",fun.aggregate=length)
kable(C_alt2)
```



## STACK

Turn the above table (C) into this (D):

| c1 | V1 | V2 |
|----|----|----|
| A  | a  | 3  |
| A  | b  | 2  |
| A  | c  | .  |
| A  | d  | .  |
| B  | a  | .  |
| B  | b  | .  |
| B  | c  | 1  |
| B  | d  | 1  |


```{r results='asis'}
D <- melt(C,id="c1") %>%
  arrange(c1)
kable(D)
```


## JOIN

Join these tables (E, F):

| c1 | c2 |
|----|----|
| A  |  1 |
| B  |  2 |
| C  |  3 |

| c1 | c3 |
|----|----|
| A  | a  |
| B  | a  |
| B  | a  |
| B  | b  |
| C  | c  |
| A  | b  |

to give (G):

| c1 | c3 | c2 |
|----|----|----|
| A  | a  |  1 |
| B  | a  |  2 |
| B  | a  |  2 |
| B  | b  |  2 |
| C  | c  |  3 |
| A  | b  |  1 |


```{r results='asis'}
E<-data.frame(c1=c("A","B","C"), c2=1:3)
FF <- data.frame(c1=c("A","B","B","B","C","A"), c3=c("a","a","a","b","c","b"))
G <- left_join(FF, E)
kable(G)
```

the `dplyr` package supplies `left_join()`, which preserves the sequence of rows in its left argument.  Alternative, as was originally suggested, one could use `merge()` :


```{r results='asis'}
G_merge <- merge(FF,E)
kable(G_merge)
```

Although columns now come out sorted.

## SUBSET 

subset Table G to give H:

| c1 | c3 | c2 |
|----|----|----|
| A  | a  |  1 |
| A  | b  |  1 |



```{r results='asis'}
H <- filter(G,c1=="A")
kable(H)
```

## TRANSPOSE

transpose H to give:

| V1 | V2 |
|----|----|
| A  | A  |
| a  | b  |
| 1  | 1  |


```{r results='asis'}
H_transpose <- data.frame(t(H))
kable(H_transpose)
```


## SORT

In the original, the question suggested "up to three keys".


```{r results='asis'}
A_arranged <- arrange(A,c1,c2,c3)
kable(A_arranged)
```


## Conclusion

To my surprise, each of these was actually a single line.  The only exception was the first (GROUP), and that was because there are really two separate steps here -- the first to actually group the data, the second to apply summary functions to the data.  `dplyr` automates both tasks, and supplies great readability.  